Active Learning for SAT Solver Benchmarking

### 2022-10-12 paper/paper.tex
Address minor comments

Let benchmark instances $\mathcal{I}$, SAT solvers $\mathcal{A}$, and runtimes $r : \mathcal{A} \times \mathcal{I} \rightarrow \left[0, \tau\right]$ be given. 
We denote the new solver to be ranked by $\hat a \not\in A$ and define $\hat A := A \cup \lbrace \hat a \rbrace$ to denote all solvers including the new solver. 
Model $\mathcal{M}$ provides a prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ that powers the decisions within our framework.

First, we initialize the model $\mathcal{M}$ in Line~1.
This creates the machine-learning prediction model and also transforms runtimes as discussed in the subsequent section.
We do not need any further steps here, since we have not obtained any labeled samples yet, so we cannot train the model yet.
The model update (Line~7) uses all acquired runtimes to train a new runtime-prediction function $f$ using common supervised machine-learning methodology.
It has also access to global instance features that assign each instance a vector in $\mathbb{R}^n$.
Note that we build a new prediction model in each iteration since running experiments (Line~5) dominates the runtime of Algorithm~\ref{algALBenchmark} by magnitudes.
Finally, we determine the solver scores that are predicted in Line~8 by using the runtime-prediction function $f$ and the approach described in the following two sections.

@@

The ranking stopping criterion is less sophisticated in comparison.
It stops the active-learning process if the ranking induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged within the last $l$ iterations.
Note that the concrete values of $s_k(\hat a)$ might still change.
We are solemnly interested in the induced ranking in this case.

@@

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{ranking}, \emph{selection}, and \emph{stopping}.
Also, there are different choices for the runtime-label prediction model and runtime discretization.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results and do the end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.

@@

\begin{itemize}
  \item Random sampling 
  \item Uncertainty sampling
  \begin{itemize}
    \item Fallback threshold: use random sampling for the first \SI{0}{\%}, \SI{5}{\%}, \SI{10}{\%}, \SI{15}{\%}, or \SI{20}{\%} of instances to explore the instance space.
    \item Runtime scaling: prefer instances with the greatest uncertainty per average runtime (over all solvers; \texttt{True} or \texttt{False}).
  \end{itemize}

  \item Information-gain sampling
  \begin{itemize}
    \item Fallback threshold: use random sampling for the first \SI{0}{\%}, \SI{5}{\%}, \SI{10}{\%}, \SI{15}{\%}, or \SI{20}{\%} of instances to explore the instance space.
    \item Runtime scaling: prefer instances with the greatest information gain per average runtime (\texttt{True} or \texttt{False}).
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
  \item Subset-size stopping criterion, using \SI{10}{\%} or \SI{20}{\%} of instances
  \item Ranking stopping criterion
  \begin{itemize}
    \item Minimum amount: sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Convergence duration: stop if the predicted ranking stays the same for a number of sampled instances equal to \SI{1}{\%} or \SI{2}{\%} of all instances.
  \end{itemize}

  \item Wilcoxon stopping criterion
  \begin{itemize}
    \item Minimum amount: sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Average of $p$-values to drop below: \SI{5}{\%}.
    \item Exponential-moving average: incorporate previous significance values by using an EMA with $\beta = 0.1$ or $\beta = 0.7$.
  \end{itemize}
\end{itemize}

@@

Our end-to-end experiments follow the evaluation framework introduced in Section~\ref{sec:exdesign:eval}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:exdesign:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
Evaluating a particular configuration with Algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
In particular, we do not show intermediate results of the active-learning procedure but only the final results after stopping.
The plotted lines represent the best performing configurations (convex hull) per top-level hyper-parameter choice, i.e., per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).

@@

Regarding the stopping criterion (Fig.~\ref{fig:annitraincolorstopping}), the ranking stopping criterion performs the most consistently well.
If accuracy is strongly favored (very high $\delta$), the Wilcoxon stopping criterion performs better.
The subset-size stopping criterion performs reasonably well but does not improve beyond a certain accuracy.

@@

Having selected the most promising hyper-parameters, we run our end-to-end active-learning experiments on the complete Anniversary Track dataset (5355 instances).
The aforementioned range $\delta \in \left[0.2, 0.8\right]$ is covered by only two configurations.
The best-performing approach for $\delta \in \left[0.2, 0.7\right]$ uses the predicted runtime-label ranking, information-gain sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with about \SI{90.48}{\%} accuracy ($O_{\operatorname{acc}}$), while only needing \SI{5.41}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).
The best-performing approach for $\delta \in (0.7, 0.8]$ uses the predicted runtime-label ranking, uncertainty sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with about \SI{92.33}{\%} accuracy ($O_{\operatorname{acc}}$), while only needing \SI{10.35}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).
These trade-offs provide a better way of prototyping and evaluating new SAT solvers rather than evaluating them on manually picked instances based on domain knowledge.

>>>>

Let benchmark instances $\mathcal{I}$, SAT solvers $\mathcal{A}$, and runtimes $r : \mathcal{A} \times \mathcal{I} \rightarrow \left[0, \tau\right]$ be given.
Model initialization (Line~1) then creates the machine-learning prediction model as well as transforms runtimes as discussed in the subsequent section.
No model training is needed at this point since we do not have any labeled data yet.
We denote the new solver to be ranked by $\hat a \not\in A$ and define $\hat A := A \cup \lbrace \hat a \rbrace$ to denote all solvers including the new solver. 
Model $\mathcal{M}$ provides a prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ that powers the decisions within our framework.
The model update (Line~7) uses all acquired runtimes to train a new runtime-prediction function $f$ using common supervised machine-learning methodology.
Training features are problem-instance features and known runtimes $r$.
After applying necessary transformations, as described in the subsequent section, we fit the machine-learning model using the transformed observations $\mathcal{R}$ as ground-truth labels.
It has also access to global instance features that assign each instance a vector in $\mathbb{R}^n$.
Note that we build a new prediction model in each iteration since running experiments (Line~5) dominates the runtime of Algorithm~\ref{algALBenchmark} by magnitudes.
Finally, we determine the solver scores that are predicted in Line~8 by using the runtime-prediction function $f$ and the approach described in the following two sections.

@@

The ranking-stopping criterion is less sophisticated in comparison.
It stops the active-learning process if the ranking induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged within the last $l$ iterations.
Note that the concrete values of $s_k(\hat a)$ might still change.
We are solemnly interested in the induced ranking in this case.

@@

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{ranking}, \emph{selection}, and \emph{stopping}.
Also, there are different choices for the runtime-label prediction model and runtime discretization.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results and do the experiments only with a subset.
The experiment configurations are given below.

@@

\begin{itemize}
  \item Random sampling 
  \item Uncertainty sampling
  \begin{itemize}
    \item Fallback threshold: Use random sampling for the first \SI{0}{\%}, \SI{5}{\%}, \SI{10}{\%}, \SI{15}{\%}, or \SI{20}{\%} of instances to explore the instance space.
    \item Runtime scaling: Whether to normalize uncertainty scores of instances by the average runtime of solvers on it or use the absolute values.
  \end{itemize}

  \item Information-gain sampling
  \begin{itemize}
    \item Fallback threshold: Use random sampling for the first \SI{0}{\%}, \SI{5}{\%}, \SI{10}{\%}, \SI{15}{\%}, or \SI{20}{\%} of instances to explore the instance space.
    \item Runtime scaling: Whether to normalize information-gain scores of instances by the average runtime of solvers on it or use the absolute values.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
  \item Subset-size stopping criterion, using \SI{10}{\%} or \SI{20}{\%} of instances
  \item Ranking stopping criterion
  \begin{itemize}
    \item Minimum amount: Sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Convergence duration: Stop if the predicted ranking stays the same for a number of sampled instances equal to \SI{1}{\%} or \SI{2}{\%} of all instances.
  \end{itemize}

  \item Wilcoxon stopping criterion
  \begin{itemize}
    \item Minimum amount: Sample at least \SI{2}{\%}, \SI{8}{\%}, \SI{10}{\%}, or \SI{12}{\%} of instances before applying the criterion.
    \item Average of $p$-values to drop below: \SI{5}{\%}.
    \item Exponential-moving average: Incorporate previous significance values by using an EMA with $\beta = 0.1$ or $\beta = 0.7$.
  \end{itemize}
\end{itemize}

@@

Our experiments follow the evaluation framework introduced in Section~\ref{sec:exdesign:eval}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:exdesign:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
Evaluating a particular configuration with Algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
In particular, we do not show intermediate results of the active-learning procedure but only the final results after stopping.
The plotted lines represent the best performing configurations (convex hull) per top-level hyper-parameter choice, i.e., per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).

@@

Regarding the stopping criterion (Fig.~\ref{fig:annitraincolorstopping}), the ranking stopping criterion performs the most consistently well.
If accuracy is strongly favored (very high $\delta$), the Wilcoxon stopping criterion performs better.
The subset-size stopping criterion performs reasonably well but does not improve beyond a certain accuracy because of sampling a fixed subset of instances.

@@

Having selected the most promising hyper-parameters, we run our active-learning experiments on the complete Anniversary Track dataset (5355 instances).
The aforementioned range $\delta \in \left[0.2, 0.8\right]$ is covered by only two configurations.
The best-performing approach for $\delta \in \left[0.2, 0.7\right]$ uses the predicted runtime-label ranking, information-gain sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with about \SI{90.48}{\%} accuracy ($O_{\operatorname{acc}}$), while only needing \SI{5.41}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).
The best-performing approach for $\delta \in (0.7, 0.8]$ uses the predicted runtime-label ranking, uncertainty sampling, and ranking stopping criterion.
It can predict a new solver's PAR-2 ranking with about \SI{92.33}{\%} accuracy ($O_{\operatorname{acc}}$), while only needing \SI{10.35}{\%} of the full evaluation time ($O_{\operatorname{rt}}$).
These trade-offs provide a better way of prototyping and evaluating new SAT solvers rather than evaluating them on manually picked instances based on domain knowledge.
